---
title: "Bike_Sharing"
output: github_document
date: "2024-08-17"
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggridges)
library(forcats)
library(corrplot)
bike_df <- read.csv("/Users/edoardochiari/Desktop/bike_sharing.csv")
```


## Changes and Info of the dataframe 

```{r}
#Removing useless variables 
bike_df<-subset(bike_df,select=-c(casual,registered))
bike_df$dteday <- as.Date( bike_df$dteday, format = "%Y-%m-%d")

#Provide the dimensions of the dataframe
dim(bike_df)

#Provide the structure of all the varibales inside the dataframe
str(bike_df)

#Provide general information (Min, Quantiles, Mean and Max) about each variable 
summary(bike_df)
```

## Visualizations

For an initial overview of the general trend of rents, a time series from 2011 to 2012 may be the best option.

```{r, message=FALSE}
ggplot(data= bike_df, aes(x= dteday, y= cnt)) + 
  geom_line(color= "#2de1c2") +
  labs(x= "Days", y= "Rents", title= "Number of rents 2011-2012") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

In order to have a better view of the trend, a line is added to provide a clearer view 

```{r,message=FALSE}
ggplot(data= bike_df, aes(x= dteday, y= cnt)) + 
  geom_line(color= "#2de1c2") +
  labs(x= "Days", y= "Rents", title= "Number of rents 2011-2012") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_smooth(color= "#FE4E00", fill= "#FFAE03", method= "loess")
```

It’s clear that bike rentals follow a seasonal pattern, with a noticeable decline in the winter months and a sharp increase during the summer. Additionally, the upward trend between 2011 and 2012 indicates a significant rise in the number of rentals over that period.


```{r}
month.abb.ita <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dic")
ggplot(data = bike_df, aes(x = mnth, y = cnt)) + 
  geom_point() + 
  labs(x = "Months", y = "Rents", title = "Average monthly rent") +
  scale_x_continuous(breaks = 1:12, labels = month.abb.ita) + 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
```

Once again, we can see that the number of rentals is significantly higher during the warmer months, particularly in September and October. This suggests a clear preference for bike-sharing in more favorable weather conditions.

To further summarize and analyze the key characteristics of the rental data, utilizing boxplots could be an effective approach. This method would allow us to visually capture the distribution, central tendency, and potential outliers within the dataset, providing deeper insights into rental patterns across different time periods.

```{r}
ggplot(data = bike_df, aes(x = mnth, y = cnt, group = mnth)) + 
  geom_boxplot(color= "#FE4E00", fill= "#2de1c2") + 
  labs(x = "Months", y = "Rents", title = "Average monthly rent") +
  scale_x_continuous(breaks = 1:12, labels = month.abb.ita) + 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
```

The median, represented by the central line in each boxplot, shows that the rental data is not symmetrically distributed since the mean and median values do not coincide. In a perfectly symmetrical distribution, the most frequent value, which splits the dataset evenly, would be the mean. To better illustrate the relationship between the mean and median, we can enhance the boxplot by adding a white dot to indicate the mean value. 

```{r}
ggplot(data = bike_df, aes(x = mnth, y = cnt, group = mnth)) + 
  geom_boxplot(color= "#FE4E00", fill= "#2de1c2") + 
  stat_summary(fun = "mean", geom = "point", col = "white", pch = 15, size = 2.5) + 
  labs(x = "Months", y = "Rents", title = "Average monthly rent - Seasonality") +
  scale_x_continuous(breaks = 1:12, labels = month.abb.ita) + 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
```

This addition suggests that bike rentals during 2011 and 2012 do not follow a symmetric distribution.

An alternative and perhaps more effective way to visualize this data is through a ridge plot. While boxplots provide a summary of the data, including medians, quartiles, and potential outliers, they don’t reveal the full shape of the data distribution. The density and ridge plots, on the other hand, show the distribution of rental counts across different months in much finer detail.

```{r, message=FALSE, warning=FALSE}
bike_df2 <- bike_df %>%
  mutate(mnth_fct = fct_relevel(factor(mnth), as.character(1:12))) %>%
  mutate(mnth_fct = fct_recode(mnth_fct, 
                               Jan = "1", Feb = "2", Mar = "3", Apr = "4", May = "5", Jun = "6", 
                               Jul = "7", Aug = "8", Sep = "9", Oct = "10", Nov = "11", Dic = "12")) %>%
  mutate(mnth_fct = fct_rev(mnth_fct)) # Reverse the order of the months

ggplot(bike_df2, aes(x = cnt, y = mnth_fct, fill = stat(x))) +
  geom_density_ridges_gradient(stat = "binline", draw_baseline = FALSE, scale = 0.95) +
  scale_fill_gradient(low = "blue", high = "red", name = "Rents") + 
  labs(y = "Months", x = "Rents", title = "Average monthly rent - Distribution") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
  
```

To improve the clarity of the graph, the histogram can be replaced with a more visually appealing format that better highlights the seasonality.

```{r, message=FALSE}
ggplot(bike_df2, aes(x = cnt, y = mnth_fct, fill = stat(x))) +
  geom_density_ridges_gradient(scale = 0.95) +
  scale_fill_gradient(low = "blue", high = "red", name = "Rents") + 
  labs(y = "Months", x = "Rents", title = "Average monthly rent - Distribution") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
```


```{r}

month_names <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                 "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

weekday_names <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")

ggplot(bike_df, aes(x = factor(mnth, levels = 1:12, labels = month_names), 
                    y = cnt, 
                    fill = factor(weekday, levels = 0:6, labels = weekday_names))) +
  geom_col() + 
  labs(x = "Months", y = "Rents", title = "Monthly Distribution of Rents by Weekday", fill= "Weekday") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_continuous(labels = scales::comma)  # Format y-axis labels as full numbers

```

```{r}
years <- c("2011", "2012")

ggplot(bike_df,aes(x=factor(yr, labels= years), y=cnt))+
  geom_violin() +
  labs(x='Years',y='Rents',title='Yearly rent distributions', fill= "Year") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# Assuming 'holidays' and 'season' are predefined vectors of labels
holiday_labels <- c("No Holiday", "Holiday")  # Replace with actual labels if different
season_labels <- c("Spring", "Summer", "Fall", "Winter")  # Replace with actual labels if different

ggplot(bike_df, aes(x = factor(holiday, levels = 0:1, labels = holiday_labels), 
                    y = cnt, 
                    fill = factor(season, levels = 1:4, labels = season_labels))) + 
  geom_col() + 
  theme_bw() +  # Consider using theme_minimal() if preferred
  labs(x = "Holiday", y = "Rents", title = "Rents Distribution by Festivity", fill = "Season") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_continuous(labels = scales::comma)  # Format y-axis labels as full numbers
```

The graph clearly indicates that the number of rentals is significantly lower during the holidays, likely due to the limited number of holidays compared to regular days.

```{r}
workingd <- c("No Working day", "Working day")

ggplot(bike_df, aes(x = factor(workingday, levels = 0:1, labels = workingd), 
                    y = cnt, 
                    fill = factor(season, levels = 1:4, labels = season_labels))) + 
  geom_col() + 
  theme_bw() +  # Consider using theme_minimal() if preferred
  labs(x = "Workingday", y = "Rents", title = "Rents Distribution by Working day", fill = "Season") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_continuous(labels = scales::comma)  # Format y-axis labels as full numbers
```

When weekends are also considered alongside holidays, the number of rentals increases. However, rentals during weekdays remain higher.

```{r}

weather <- c("Clear sky", "Cloudy sky", "Light precipitation", "Heavy precipitation")

ggplot(bike_df, aes(x= factor(weathersit, levels = 1:4, labels = weather),
       y= cnt,
       fill = factor(season, levels = 1:4, labels = season_labels))) +
  geom_col() + 
  theme_bw() +  # Consider using theme_minimal() if preferred
  labs(x = "Weather", y = "Rents", title = "Rents Distribution by type of Weather", fill = "Season") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_continuous(labels = scales::comma)  # Format y-axis labels as full numbers
```

The graph reveals that the majority of bike-sharing activity occurs during favorable weather conditions, with clear skies being the most common. This is followed by periods of overcast skies, where bike usage slightly declines but remains significant. Finally, the graph indicates that light precipitation leads to the lowest levels of bike-sharing, as riders tend to avoid cycling in wet conditions. This pattern highlights the strong influence of weather on biking habits, with people preferring to ride when the weather is more comfortable and avoiding it when conditions are less favorable.

While the weather graphs provided a broad overview of how general conditions affect bike usage, they did not fully capture the nuanced impact of specific factors like temperature, humidity, and wind speed.
To gain deeper insights into these variables, a boxplot analysis can be conducted. This method allowes for a more precise examination of the distribution and variability of each weather factor, as well as the identification and removal of outliers. By doing so, the study could focus on the typical range of conditions that most significantly influence bike-sharing behavior. 

### Outliers for temperature, humidity and wind speed:

```{r}
#Insert the boxplot in three different columns in one row
par(mfrow=c(1,3)) 

#Box plot for temoperature
boxplot(bike_df$temp, main="Temperature",col="plum1",horizontal=F)
  
#Box plot for humidity
boxplot(bike_df$hum,main="Humidity",col="cadetblue1",horizontal=F)

#Box plot for wind speed 
boxplot(bike_df$windspeed,main="Wind Speed",col="darkolivegreen3",horizontal=F)
```

Since the temperature values doesn't present outliers, only humidity and wind speed require the removal

Outliers humidity
```{r}
QH <- quantile(bike_df$hum, probs=c(.25, .75), na.rm = T)
iqrH <- IQR(bike_df$hum, na.rm = T)

# remove the outlier beyond 1.5 * iqr for Rentals
bike_sharing2 <- bike_df %>% filter(hum > (QH[1] - 1.5*iqrH) & 
                                      hum < (QH[2]+1.5*iqrH))  

# visualize the new dataset without outliers
par(mfrow=c(2,1))
options(repr.plot.width=10, repr.plot.height=7)
boxplot(bike_df$hum, col = "cadetblue1", main = "Humidity before outliers removal", horizontal = T) 
boxplot(bike_sharing2$hum, col = "cadetblue1", main = "Humidity after outliers removal", horizontal = T)
```

Outliers windspeed
```{r}
# Calculate quantiles and IQR for windspeed
QW <- quantile(bike_df$windspeed, probs = c(.25, .75), na.rm = T)
iqrW <- IQR(bike_df$windspeed, na.rm = T)

# Filter out the outliers for windspeed
bike_sharing2 <- bike_df %>%
  filter(windspeed > (QW[1] - 1.25 * iqrW) & 
         windspeed < (QW[2] + 1.25 * iqrW))

# Visualize the result
par(mfrow = c(2,1))
options(repr.plot.width = 10, repr.plot.height = 7)
boxplot(bike_df$windspeed, col = "darkolivegreen3", main = "Windspeed before outliers removal", horizontal = T) 
boxplot(bike_sharing2$windspeed, col = "darkolivegreen3", main = "Windspeed after outliers removal", horizontal = T)
```

The removal of outliers is particularly important to ensure that the analysis focused on normal conditions rather than being skewed by rare, extreme weather events, which could distort the findings. For instance, an outlier in wind speed might represent an unusually strong storm that, while interesting, does not reflect the usual conditions under which most bike-sharing occurs. By excluding such outliers, the study could more accurately capture the relationship between these variables and bike usage under typical circumstances.

Furthermore, this approach enhances the reliability and accuracy of any predictive models developed as part of the study. Outliers can introduce noise and reduce the model’s ability to generalize to everyday conditions, leading to less accurate predictions. By removing these outliers, the analysis and any resulting models are more likely to produce insights that are both relevant and actionable for decision-makers, such as bike-sharing companies or urban planners. Overall, this method ensures that the study’s conclusions are grounded in the data that most accurately represents regular biking conditions.


## Correlations

Let's investigate the correlation of the different variables in the dataframe
```{r}
#In order to get the correlation, all interested variables need to be converted into numeric
bike_df$yr<-as.numeric(bike_df$yr)
bike_df$mnth<-as.numeric(bike_df$mnth)
bike_df$season <- as.numeric(bike_df$season)
bike_df$holiday<- as.numeric(bike_df$holiday)
bike_df$weekday<- as.numeric(bike_df$weekday)
bike_df$workingday<- as.numeric(bike_df$workingday)
bike_df$weathersit<- as.numeric(bike_df$weathersit)

#Correlation
data <- bike_df %>% 
  select(season, temp, atemp, weathersit, mnth, yr, holiday, weekday, workingday,holiday ) %>%
  drop_na()
round(cov(data), 2)

#Correlation plot 
corrplot(corr= cor(bike_df[3:14]),
         method = "circle",
         type = "full",
         tl.pos = "tl",
         order= "original")

```

The covariance matrix reveals several insights into the relationships between the variables. A notable observation is the strong positive covariance between the season and month variables, which reflects the natural progression of months through different seasons. This relationship is expected, given that the month of the year directly influences the season.

When examining the weather situation in relation to temperature and apparent temperature, a slight negative covariance is observed. This suggests that as weather conditions worsen (for example, during rain or snow), temperatures tend to be slightly lower, though this relationship appears to be relatively weak.

Additionally, the positive covariance between the month and both temperature and apparent temperature indicates that as the year progresses, particularly moving from winter to summer, temperatures generally increase. This is consistent with typical seasonal weather patterns.

Other relationships in the matrix, such as those involving holidays, weekdays, and working days, show little to no significant covariance with the other variables. These minimal values suggest that holidays and weekdays do not strongly covary with other factors like temperature, season, or weather conditions in this dataset, indicating a more independent relationship.
